---
title: "XtX subpopulation differentiation"
output:
  html_document:
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 4
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 4
  word_document: default
date: '`r format(Sys.Date(),"%d/%b/%Y")`'
---

```{r knitr setup, include=FALSE,  eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
knitr::opts_chunk$set(eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, 
                      comment = "", results="markup")
```
This document describes how SNP genotypes from 135 landraces of the Spanish Barley Core Collection (SBCC) grouped into 4 subpopulations, were used to obtain XtX statistics by CP Cantalapiedra.


# Rebuilding data files to obtain subpopulation-based bayenv input files

## Files for covariance matrix

```{r b1, engine='bash'}
cd xtx_subpops
./bayenv_lines_to_subpops.py SBCC_order.subpops.tab ../SBCC_order.txt ../SBCC_nr_SNPs.tsv show \
> SBCC_nr_subpops.tsv 2> SBCC_nr_subpops.err
```

## Files for bayenv XtX analysis

```{r b2, engine='bash'}
cd xtx_subpops
./bayenv_lines_to_subpops.py SBCC_order.subpops.tab ../SBCC_order.txt ../SBCC_9K_SNPs.tsv show \
> SBCC_9K_subpops.tsv 2> SBCC_9K_subpops.err
```

**NOTE**: the "show" option is used to include in the output file those SNPs with missing data in some of the genotypes. The opposite could be obtained using "hide" as option.

# Running bayenv2 XtX analyses

Following the work of [Rusell et al 2016](http://www.nature.com/ng/journal/v48/n9/full/ng.3612.html), 
200K MCMC steps are performed.

First, a dummy climate file is required in order to run bayenv:

```{r b3, engine='bash'}
cd xtx_subpops
cat ../SBCC_environfile.tsv | head -1 | awk '{print $1"\t"$2"\t"$3"\t"$4}' > envfile.dummy
```

A covariance matrix is also required. Here it is computed 10 times:

```{r b4, engine='bash', eval=FALSE}
cd xtx_subpops
mkdir -p matrices/raw/

# Create 10 covariance matrices, this will take some time
for i in {1..10}; do
	rnd=$(perl -e 'printf("%05d",rand(99999))');
	echo $rnd >> bayenv.covar.rnd_seeds
	mkdir _job$i; cd _job$i; ln -s ../SBCC_nr_subpops.tsv .; \
	../soft/bayenv2/bayenv2 -i SBCC_nr_subpops.tsv -p 4 -k 100000 -r $rnd \
	> ../SBCC_nr_subpops_matrix_it100K_$i.out & 
	cd ..;
done;

# We can now put the resulting files away:
rm -rf _job*
gzip SBCC_nr_subpops_matrix_it100K_*
mv SBCC_nr_subpops_matrix_it100K_* matrices/raw/

# Extract final matrix from each replicate
for i in {1..10}; do 
  zcat matrices/raw/SBCC_nr_subpops_matrix_it100K_$i.out.gz | \
  perl -lne 'if(/ITER = 100000/){$ok=1}elsif($ok){ print }' \
  | sed '/^$/d' > matrices/SBCC_nr_subpops_matrix_$i.txt
done
```
A single average matrix can be computed as follows:

```{r average, eval=FALSE}
setwd(xtx_subpops)

# read all final matrices
m1 = as.matrix( read.table(file="matrices/raw/SBCC_nr_subpops_matrix_1.txt", header=F) )
m2 = as.matrix( read.table(file="matrices/SBCC_nr_subpops_matrix_2.txt", header=F) )
m3 = as.matrix( read.table(file="matrices/SBCC_nr_subpops_matrix_3.txt", header=F) )
m4 = as.matrix( read.table(file="matrices/SBCC_nr_subpops_matrix_4.txt", header=F) )
m5 = as.matrix( read.table(file="matrices/SBCC_nr_subpops_matrix_5.txt", header=F) )
m6 = as.matrix( read.table(file="matrices/SBCC_nr_subpops_matrix_6.txt", header=F) )
m7 = as.matrix( read.table(file="matrices/SBCC_nr_subpops_matrix_7.txt", header=F) )
m8 = as.matrix( read.table(file="matrices/SBCC_nr_subpops_matrix_8.txt", header=F) )
m9 = as.matrix( read.table(file="matrices/SBCC_nr_subpops_matrix_9.txt", header=F) )
m10 = as.matrix( read.table(file="matrices/SBCC_nr_subpops_matrix_10.txt", header=F) )

# make a list of matrices and get mean as explained in:
mat_list = list( m1, m2, m3, m4, m5, m6, m7, m8, m9, m10 )
mean_mat = apply(simplify2array(mat_list), c(1,2), mean)

# write resulting mean cov matrix
write.table(mean_mat,file="matrices/SBCC_nr_subpops_matrix_mean.txt",
            sep="\t",row.names=F,col.names=F,quote=F)
```

This should create file *xtx_subpops/matrices/SBCC_nr_subpops_matrix_mean.txt*

Now three replicates of bayenv XTX will be run:

```{r b5, engine='bash', eval=FALSE}
cd xtx_subpops

# 1st replicate run
rnd=$(perl -e 'printf("%05d",rand(99999))');
echo "$rnd" > SBCC_9K_subpops.tsv.xtx.rep1.rnd

../soft/bayenv2/calc_xtx_parallel.pl 5 bayenv2 -X -t -i SBCC_9K_subpops.tsv -p 4 \
-e envfile.dummy -n 1 \
-m matrices/SBCC_nr_subpops_matrix_mean.txt \
-k 200000 -r "$rnd" -o SBCC_9K_subpops.tsv.xtx.rep1 -c \
> SBCC_9K_subpops.tsv.xtx.rep1.stdout 2>&1;

# 2nd run
rnd=$(perl -e 'printf("%05d",rand(99999))');
echo "$rnd" > SBCC_9K_subpops.tsv.xtx.rep2.rnd

../soft/bayenv2/calc_xtx_parallel.pl 5 bayenv2 -X -t -i SBCC_9K_subpops.tsv -p 4 \
-e envfile.dummy -n 1 \
-m matrices/SBCC_nr_subpops_matrix_mean.txt \
-k 200000 -r "$rnd" -o SBCC_9K_subpops.tsv.xtx.rep2 -c \
> SBCC_9K_subpops.tsv.xtx.rep2.stdout 2>&1;

# 3rd run
rnd=$(perl -e 'printf("%05d",rand(99999))');
echo "$rnd" > SBCC_9K_subpops.tsv.xtx.rep3.rnd

../soft/bayenv2/calc_xtx_parallel.pl 5 bayenv2 -X -t -i SBCC_9K_subpops.tsv -p 4 \
-e envfile.dummy -n 1 \
-m matrices/SBCC_nr_subpops_matrix_mean.txt \
-k 200000 -r "$rnd" -o SBCC_9K_subpops.tsv.xtx.rep3 -c \
> SBCC_9K_subpops.tsv.xtx.rep3.stdout 2>&1;

# compress outfiles
gzip SBCC_9K_subpops.tsv.xtx.rep1
gzip SBCC_9K_subpops.tsv.xtx.rep2
gzip SBCC_9K_subpops.tsv.xtx.rep3
```

Let's inspect one of the XtX bayenv files:

```{r, engine='bash'}
cd xtx_subpops
zcat SBCC_9K_subpops.tsv.xtx.rep1.gz | head -6
```


# Annotating and exporting XtX results 

```{r, engine='bash'}
cd xtx_subpops

# add SNP name to XtX results
perl xtx2table.pl SBCC_9K_subpops.tsv.xtx.rep1.gz ../SBCC_9K_SNPs.annot.tsv > \
  SBCC_9K_subpops.tsv.xtx.rep1.tsv
perl xtx2table.pl SBCC_9K_subpops.tsv.xtx.rep2.gz ../SBCC_9K_SNPs.annot.tsv > \
  SBCC_9K_subpops.tsv.xtx.rep2.tsv
perl xtx2table.pl SBCC_9K_subpops.tsv.xtx.rep3.gz ../SBCC_9K_SNPs.annot.tsv > \
  SBCC_9K_subpops.tsv.xtx.rep3.tsv

head -6 SBCC_9K_subpops.tsv.xtx.rep1.tsv

gzip -f SBCC_9K_subpops.tsv.xtx.rep1.tsv
gzip -f SBCC_9K_subpops.tsv.xtx.rep2.tsv
gzip -f SBCC_9K_subpops.tsv.xtx.rep3.tsv


# add cM & bp map positions (./obtain_map.R IN OUT MAP)

Rscript ./obtain_map.R SBCC_9K_subpops.tsv.xtx.rep1.tsv.gz \
  SBCC_9K_subpops.tsv.xtx.rep1.bp.tsv ../raw/9920_SNPs_SBCC_bp_map2017.curated.tsv 
Rscript ./obtain_map.R SBCC_9K_subpops.tsv.xtx.rep2.tsv.gz \
  SBCC_9K_subpops.tsv.xtx.rep2.bp.tsv ../raw/9920_SNPs_SBCC_bp_map2017.curated.tsv 
Rscript ./obtain_map.R SBCC_9K_subpops.tsv.xtx.rep3.tsv.gz \
  SBCC_9K_subpops.tsv.xtx.rep3.bp.tsv ../raw/9920_SNPs_SBCC_bp_map2017.curated.tsv 

Rscript ./obtain_map.R SBCC_9K_subpops.tsv.xtx.rep1.tsv.gz \
  SBCC_9K_subpops.tsv.xtx.rep1.cM.tsv ../raw/9920_SNPs_SBCC_cM_map2017.curated.tsv 
Rscript ./obtain_map.R SBCC_9K_subpops.tsv.xtx.rep2.tsv.gz \
  SBCC_9K_subpops.tsv.xtx.rep2.cM.tsv ../raw/9920_SNPs_SBCC_cM_map2017.curated.tsv 
Rscript ./obtain_map.R SBCC_9K_subpops.tsv.xtx.rep3.tsv.gz \
  SBCC_9K_subpops.tsv.xtx.rep3.cM.tsv ../raw/9920_SNPs_SBCC_cM_map2017.curated.tsv

# join all results into a single file
join -t $'\t' SBCC_9K_subpops.tsv.xtx.rep1.bp.tsv SBCC_9K_subpops.tsv.xtx.rep2.bp.tsv \
  | join -t $'\t' - SBCC_9K_subpops.tsv.xtx.rep3.bp.tsv \
  | awk '{print $1"\t"$3"\t"$4"\t"$2"\t"$5"\t"$8}' \
  >  SBCC_9K_subpops.tsv.xtx.bp.tsv 

join -t $'\t' SBCC_9K_subpops.tsv.xtx.rep1.cM.tsv SBCC_9K_subpops.tsv.xtx.rep2.cM.tsv \
  | join -t $'\t' - SBCC_9K_subpops.tsv.xtx.rep3.cM.tsv \
  | awk '{print $1"\t"$3"\t"$4"\t"$2"\t"$5"\t"$8}' \
  >  SBCC_9K_subpops.tsv.xtx.cM.tsv

head -6 SBCC_9K_subpops.tsv.xtx.bp.tsv

head -6 SBCC_9K_subpops.tsv.xtx.cM.tsv
```

The final results file of tis protocol are
[SBCC_9K_subpops.tsv.xtx.bp.tsv](xtx_subpops/SBCC_9K_subpops.tsv.xtx.bp.tsv) and
[SBCC_9K_subpops.tsv.xtx.cM.tsv](xtx_subpops/SBCC_9K_subpops.tsv.xtx.cM.tsv)

