---
title: "Redundancy Analysis (RDA)"
output:
  html_document:
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 4
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 4
  word_document: default
date: '`r format(Sys.Date(),"%d/%b/%Y")`'
---

```{r knitr setup, include=FALSE,  eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
knitr::opts_chunk$set(eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, 
                      comment = "", results="markup")
```

This file explains the procedure followed by Ernesto Igartua to perform redundancy analysis on genetic features of the genotypes (probability of belonging to each one of the four germplasm groups), explained by geographic and agro-climatic features of the locations of collection of the genotypes. This uses library **vegan**. 


```{r}
library(vegan)
```

Input of environmental, geographic, genetic data
```{r data input}

#Input data, (Q matrix, all geographic and agro-climatic variables for all accessions)
SBCC <- read.table("RDA/RDA_complete.tsv", header = TRUE)

#Probabilities derived from structure, 4 populations or germplasm groups
Cluster <- SBCC[,c("Q1","Q2","Q3","Q4")]

#Dummy variables
Dummies <- SBCC[,6:17]

#All explanatory variables, environmental, geographic, dummies
EnvGeoDum<-SBCC[,6:ncol(SBCC)]

#Just geographic variables
Geo <- SBCC[,c("altitude","utmx","utmy")]

#Just environmental variables, all, non-standardized
Env <- SBCC[,21:ncol(SBCC)]
head(SBCC)

#Input data for a third redundancy analysis. Input of set of variables chosen by cluster analysis, 17 agro-climatic, plus 3 geographic and 12 dummies; they are already standardized

#Input data, (Q matrix, all geographic and agro-climatic variables for all accessions)
SBCCc<- read.table("RDA/RDA3.tsv", header = TRUE)

#All variables, geographical, dummies, environmental, all standardized
Allc<-SBCCc[,c("dummy_01","dummy_02","dummy_03","dummy_04","dummy_05","dummy_06","dummy_07",
             "dummy_08","dummy_09","dummy_10","dummy_11","dummy_12","lon","lat","alt",
             "verna_30d","verna_jan_feb","verna_mar_apr","pfrost_01","pcp_aut","pcp_win",
             "pcp_mar_apr","pcp_may_jun","frost_jan_feb","frost_apr_may","tamp_win",
             "tamp_spr","et0_spr","bal_aut","bal_win","bal_jun","bal_mar_apr_may")]

#Just dummies
Dummyc<-SBCCc[,c("dummy_01","dummy_02","dummy_03","dummy_04","dummy_05","dummy_06","dummy_07",
             "dummy_08","dummy_09","dummy_10","dummy_11","dummy_12")]

#Just geographic
Geoc<-SBCCc[,c("lon","lat","alt")]

#Just environmental (agro-climatic)
Envc<-SBCCc[,c("verna_30d","verna_jan_feb","verna_mar_apr","pfrost_01","pcp_aut","pcp_win",
             "pcp_mar_apr","pcp_may_jun","frost_jan_feb","frost_apr_may","tamp_win",
             "tamp_spr","et0_spr","bal_aut","bal_win","bal_jun","bal_mar_apr_may")]

#Just probabilities derived from structure, 4 populations or germplasm groups
Clusterc <- SBCCc[,c("Q1","Q2","Q3","Q4")]
```

Standardization of variables 
```{r}
#Standardization of variables
scaled.EnvGeoDum<-scale(EnvGeoDum)
scaledEnvGeoDum.df<-as.data.frame(scaled.EnvGeoDum)

scaled.env<-scale(Env)
scaledenv.df <- as.data.frame(scaled.env)

scaled.geo<-scale(Geo)
scaledgeo.df <- as.data.frame(scaled.geo)
```

First redundancy analysis with just environmental (agroclimatic) variables. Then selection of most significant variables is done with step-wise regression, and a second, reduced, redundancy analysis is performed, with just the variables selected in multiple regression with default settings, until the first dummy variable was entered into the model. 
```{r}
#Redundancy analysis, complete data set
SBCC_RDA<-rda(Cluster,scaledenv.df,scale=F)

#Stepwise selection of variables, complete data, environmental, geographic and dummy variables
RDA1<-rda(Cluster~1,scaledEnvGeoDum.df)
RDA2<-rda(Cluster~.,scaledEnvGeoDum.df)
step.forward<-ordistep(RDA1,scope=formula(RDA2), direction="forward",perm.max=200,pstep=999)

#Redundancy analysis, complete data reduced to env variables selected by forward selection, until first dummy variable entered in the model
SBCC_RDA_red<-rda(Cluster~pfrost_01+bal_06, data=scaledenv.df)
```

Information on outcome of RDA. Variance inflation ratio informs on multicolinearity (values above 10 are too high). R-squared adjusted for the number of parameters indicates proportion of variation explained
```{r}
#variance inflation ratio, above 10 indicates multicolinearity
vif.cca(SBCC_RDA)
vif.cca(SBCC_RDA_red)

#R-squared, raw and adjusted, explained by rda
RsquareAdj(SBCC_RDA_red)
RsquareAdj(SBCC_RDA)
```

Plots of both redundancy analyses, scaling=0 chosen, as it results in good visualization of the plot; see help of the package for other scaling options
```{r}
#Triplot of rda results, check scaling options in help file
plot(SBCC_RDA,type="text", scaling=0)
plot(SBCC_RDA_red,type="text", scaling=0)
```

Third redundancy analysis with just 17 chosen environmental (agroclimatic) variables.  
```{r}
#Redundancy analysis, complete data set
SBCCc_RDA<-rda(Clusterc,Envc,scale=F)
```

Information on outcome of RDA. Variance inflation ratio informs on multicollinearity (values above 10 are too high). R-squared adjusted for the number of parameters indicates proportion of variation explained
```{r}
#variance inflation ratio, above 10 indicates multicolinearity
vif.cca(SBCCc_RDA)

#R-squared, raw and adjusted, explained by rda
RsquareAdj(SBCCc_RDA)
```

Plot of redundancy analyses, scaling=0 chosen, as it results in good visualization of the plot; see help of the package for other scaling options
```{r}
#Triplot of rda results, check scaling options in help file
plot(SBCCc_RDA,type="text", scaling=0)
```

Variance of distribution of germplasm groups explained by #1 full model (all agro-climatic variables); #2 geographic variables, after agro-climatic variables are taken into account; #3 agro-climatic variables are taken into account. Significance calculated. 
```{r}
#Variance explained by different fractions of variables
#1
anova(SBCCc_RDA)
#2
anova.cca(rda(Clusterc,Geoc,Envc),step=1000)
#3
anova.cca(rda(Clusterc,Envc,Geoc),step=1000)
```

Percentage of variation of distribution of germplasm groups explained by environmental (agro-climatic) variables alone (fraction a), geographic variables alone (fraction c), environmental and geographic simultaneously (fraction b). Self-explanatory plot
```{r}
# Variation partitioning
spe.part <- varpart(Clusterc, Envc, Geoc)
spe.part
plot(spe.part, digits=2)
```

Test of variance explained by different sets of variables
```{r}
# Test of fractions [a+b], only environmental variables
anova.cca(rda(Clusterc, Envc), step=1000)
# Test of fractions [b+c], only geographic variables
anova.cca(rda(Clusterc, Geoc), step=1000)
# Test of fractions [a+b+c], union of environmental and geographic
env.pars <- cbind(Envc, Geoc)
anova.cca(rda(Clusterc, env.pars), step=1000)
#Test of fraction [a], environmental variables, excluding variation shared with geographic
anova.cca(rda(Clusterc, Envc, Geoc), step=1000)
#Test of fraction [c], geographic variables, excluding variation shared with environmental (b)
anova.cca(rda(Clusterc, Geoc, Envc), step=1000)
```

Export results, files stored in directory RDA:
```{r}
Scores<-scores(SBCCc_RDA)
Summaries<-summary(SBCCc_RDA)
write.table(Scores$sites,'RDA/scores_genotypes.txt', sep='\t')
write.table(Summaries$biplot,'RDA/scores_env.txt', sep='\t')
write.table(Scores$species,'RDA/scores_germplasmgroups.txt', sep='\t')
```
